\documentclass[11pt]{article}

\usepackage[review]{acl}

% Standard package includes from the ACL LaTeX template.
\usepackage{times}
\usepackage{latexsym}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{url}
\usepackage{microtype}
\usepackage{inconsolata}
\usepackage{graphicx}

\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{listings}

\title{Intent-Compiled Verified Execution for Reliable Stateful Tool Use}
\author{}

\newcommand{\icve}{ICVE\xspace}
\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  xleftmargin=0pt,
  xrightmargin=0pt
}

\begin{document}
\maketitle

\begin{abstract}
Language agents increasingly operate state-mutating tools, where one wrong argument or
repeated action can cause invalid calls, unsafe state changes, and costly repair loops.
We study \textbf{Intent-Compiled Verified Execution (ICVE)}, a runtime-checked framework
for covered high-risk intents. ICVE compiles a request into a typed intent frame and
delegates evidence acquisition, slot grounding, precondition repair, abstention,
execution, and postcondition checking to deterministic state machines. The model still
identifies intent, but no longer owns mutating control flow for registered task families.
On public ToolSandbox single-turn and insufficient-information tasks, ICVE changes the
failure mode: covered mutating requests become checked transitions or guarded abstentions
with zero invalid or unsafe calls, while using far fewer model calls and tokens than
ReAct. The pattern holds across local Qwen2.5 scales, a Phi-3-mini check, and hosted
DeepSeek replications. AppWorld evaluations show the same coverage-risk profile on the
official dev split and local \texttt{test\_normal.txt}, while uncovered
\texttt{test\_challenge} families are reported as guarded no-action outcomes rather than
forced mutations.
\end{abstract}

\section{Introduction}

Stateful tool use differs from stateless API answering. A single wrong argument can delete
the wrong contact, send a message under the wrong precondition, or repeat an already
completed state transition until the execution environment raises an error. Existing
tool-use agents often represent this entire process as a text loop: the language model
chooses a tool, fills arguments, interprets observations, repairs errors, and decides when
to stop. ReAct-style agents, tool-learning systems, and recent interactive benchmarks
make this setting increasingly realistic \citep{yao2022react,schick2023toolformer,
lu2024toolsandbox,appworld2024,wang2025dialogtool}, but they leave open where stateful
execution should be verified.

Our thesis is that many failures are not reasoning failures alone. They are
\emph{execution-boundary} failures: the model is asked to perform intent compilation,
state tracking, evidence lookup, precondition management, mutation, and completion
detection through plain text. Larger models help, but they do not remove invalid calls or
unsafe state changes in insufficient-information settings.

ICVE changes that boundary. The LLM may still handle uncovered or low-risk requests, but
registered state-sensitive intents are routed through a runtime compiler that emits
runtime-checked tool traces. The reusable unit is not a prompt template but a declared
intent machine whose runtime owns slots, evidence, repair, abstention, transitions, and
postcondition stopping. The headline evaluation uses manually declared machines; the
artifact also includes a restricted affordance-induction prototype for regular boolean
setting APIs. The claim is an explicit, auditable boundary for covered task families and
signature-identifiable affordances, not general machine synthesis or universal tool
safety.

The positioning is deliberately narrow. ICVE is not merely stricter schema validation: a
function schema can reject an ill-typed call, but it does not decide which evidence must
be acquired before a mutation, whether an earlier transition already satisfied the goal,
or whether missing state should force abstention. Nor is ICVE only semantic parsing from
utterances to API calls: the compiled object is an executable state machine whose
transition is checked against a live ledger. Unlike proof-carrying prompting, recurring
proof obligations are encoded in the registered machine rather than regenerated by the
model. Throughout the paper, ``verified'' means runtime checking against declared machine
invariants, not a formal proof of arbitrary agent behavior.

This paper makes three contributions:
\begin{enumerate}[leftmargin=*]
\item A typed intent/state-machine abstraction for stateful tool use. Each risky task
family is represented as an \texttt{IntentMachine} with an explicit schema, compiler, and
handler.
\item A runtime-checked executor that owns evidence acquisition, argument grounding,
precondition repair, insufficient-information abstention, and completion detection.
\item A public-benchmark evaluation on ToolSandbox and AppWorld showing a failure-mode
shift from invalid or unsafe mutation toward checked transitions and guarded abstention on
covered families, plus a bounded no-LLM diagnostic showing that regular boolean-setting
affordances can be induced from previously unseen API signatures.
\end{enumerate}

\begin{figure}[t]
\centering
\scriptsize
\setlength{\fboxsep}{0.7pt}
\newcommand{\boundbox}[3]{\fcolorbox{black!45}{#1}{\parbox[c][0.34cm][c]{#2}{\centering #3}}}
\begin{tabular}{@{}p{0.44\columnwidth}@{\hspace{0.012\columnwidth}}p{0.44\columnwidth}@{}}
\multicolumn{1}{c}{\textbf{Free-form loop}} &
\multicolumn{1}{c}{\textbf{\icve state machine}}\\[-0.15em]
\fcolorbox{red!40}{red!4}{\parbox[t][2.23cm][t]{0.415\columnwidth}{
\centering
\textit{model owns search, mutate, stop}\\[-0.1em]
\boundbox{red!9}{0.33\columnwidth}{guess evidence / args}\\[-0.05em]
\(\downarrow\)\\[-0.45em]
\boundbox{red!9}{0.33\columnwidth}{mutating tool call}\\[-0.05em]
\(\downarrow\)\\[-0.45em]
\boundbox{red!9}{0.33\columnwidth}{parse error and retry}\\[-0.05em]
\(\leftarrow\) unsafe loop
}} &
\fcolorbox{blue!40}{blue!3}{\parbox[t][2.23cm][t]{0.415\columnwidth}{
\centering
\textit{model-owned}\\[-0.1em]
\boundbox{blue!8}{0.33\columnwidth}{request \(\to\) IntentFrame}\\[-0.05em]
\(\Downarrow\) \textit{boundary shift}\\[-0.35em]
\textit{runtime-owned}\\[-0.1em]
\boundbox{green!9}{0.33\columnwidth}{schema + ledger}\\[-0.05em]
\(\downarrow\)\\[-0.45em]
\boundbox{green!9}{0.33\columnwidth}{repair or abstain}\\[-0.05em]
\(\downarrow\)\\[-0.45em]
\boundbox{green!9}{0.33\columnwidth}{action + stop}
}}
\end{tabular}
\caption{The execution-boundary shift. A free-form agent owns evidence search, mutation,
error recovery, and stopping, so missing state can become invalid calls or unsafe loops.
\icve restricts the model to typed frame selection and moves ledger checks, repair,
abstention, tool execution, and postcondition stopping into the runtime.}
\label{fig:icve-overview}
\end{figure}

\section{Method}

\subsection{Typed Intent Machines}

ICVE represents each covered stateful task family as an intent machine:
\begin{lstlisting}
Machine  = (schema, compiler, handler)
schema   = (intent_type, slots, terminal_states)
compiler = request x dialogue x tools -> frame or None
handler  = frame x ledger x tools -> tool_action or final_action
\end{lstlisting}

An \texttt{IntentFrame} is the runtime-owned state for a user intent:
\begin{lstlisting}
IntentFrame = (
  intent_type,
  slots: name -> (value, required, source),
  state,
  missing_slots,
  abstain_reason
)
\end{lstlisting}

The compiler is deliberately typed and conservative. It does not ask the model to emit a
proof or a full tool trace. Instead, it maps a user request into a small frame whose slot
values are extracted from the request, copied from prior observations, or produced by a
canonicalizer. In the ToolSandbox binding, the registry currently covers 13 intent
families: currency conversion, stock lookup, reverse geocode, current city, last-message
contact update, bulk contact relationship update, contact name lookup, location phone
lookup, holiday timestamp/countdown, weather, distance, and reminder.

The machine interface is also the method's trust boundary. A model may choose the wrong
intent type or slot; ICVE mitigates that risk with schema validation, conservative
compilers, domain verifiers, and abstention. For covered intents, the model cannot choose
arbitrary state-changing calls: handlers emit only actions derived from validated slots
and ledger observations, and a policy verifier can reject or rewrite the action. Thus the
claim is conditional and inspectable: behavior depends on the registered machine and
verifier, not on an opaque promise that the model will reason safely.

\subsection{Compilation}

Compilation proceeds by registry order. Figure~\ref{fig:icve-overview} illustrates the
resulting boundary: the compiler selects a frame, while the runtime owns state checks,
repair, abstention, execution, and stopping.
On each turn, ICVE first runs an insufficient-information verifier. If required tools or
information are unavailable, the verifier returns an abstention frame. Otherwise, the
runtime tries registered compilers until one returns a frame, and validates that frame
against its schema. If no registered machine applies, control falls back only for
uncovered low-risk requests.

This loop is implemented in a benchmark-independent runtime named \texttt{RaveRuntime}
in the artifact for historical reasons. Domain bindings provide intent machines, a
state-ledger interface, and a verifier for final-message wording and transition
rejection. The generic runtime owns registry lookup, uniqueness checks, frame validation,
completion/abstention hooks, and handler dispatch. A policy verifier can rewrite a
\texttt{ToolAction} into a \texttt{FinalAction}, giving each domain binding one explicit
boundary for unsafe, unsupported, or previously failed transitions.

The artifact also contains an intentionally narrow affordance-template induction path.
When enabled, the runtime records unsupported requests, discovers regular boolean
setting affordances from tool names/signatures of the form \verb|get_*_status()| and
\verb|set_*_status(on: bool)|, proposes a candidate \texttt{IntentMachine}, executes it
in shadow mode, and promotes it only if tool-signature invariants, emitted-action
whitelists, and unrelated-request counterexamples pass. This dynamic path is a coverage
diagnostic for signature-identifiable stateful affordances, not evidence that arbitrary
state machines can be synthesized.

\subsection{Runtime-Checked State Machines}

The runtime consumes a state ledger rather than trusting raw dialogue text. The ledger
records user-provided values, successful tool observations, and recent tool errors; a
ToolSandbox ledger instantiates this interface with settings, contacts, messages, and
reminders. A state-machine handler consumes the current \texttt{IntentFrame} and ledger,
then emits at most one transition:
\begin{verbatim}
ToolAction  = (tool, args, reason)
FinalAction = (message, reason)
\end{verbatim}

For a registered machine, a mutating \texttt{ToolAction} is admissible only when the
frame is schema-valid, required slots are grounded, preconditions hold or a bounded
repair is emitted, the postcondition is not already satisfied, and the action belongs to
the handler's transition set. Otherwise the runtime must stop, abstain, or route to a
low-risk fallback. This is the measured runtime invariant, not a proof about arbitrary
unregistered behavior.

For read-only actions, the verifier normalizes arguments and blocks empty or ungrounded
searches. For state-changing actions, it checks schema conformance, argument grounding,
state preconditions, and duplicate completion before allowing exactly one transition. If
the requested postcondition is already satisfied, the runtime stops. When a safe repair
exists, such as disabling low-battery mode before enabling Wi-Fi, the runtime emits the
repair action before the requested mutation. When repair would require missing
information or unavailable tools, it abstains instead of letting the model guess.
Thus the verifier is stricter than a call schema: it checks the transition's evidence,
preconditions, and completion state, not only the argument types of one invocation.

The design point is that stateful execution is runtime-owned: registered intents are
compiled and executed through deterministic machines whose assumptions are visible in
code, while uncovered or low-risk requests can still use the model directly.

\section{Experimental Setup}

We evaluate under real local LLM inference rather than oracle policy traces. The main
agent LLM is served through a local OpenAI-compatible CUDA server. ToolSandbox tools that
would otherwise require external RapidAPI calls use deterministic offline fixtures so the
benchmark remains repeatable.

All reported local-model experiments use temperature 0 inference on a single NVIDIA RTX
3080 Ti GPU with 12GB memory. Qwen2.5-7B and Phi-3-mini are run with 4-bit quantization;
Qwen2.5-0.5B and Qwen2.5-3B are run as local instruction models through the same
OpenAI-compatible interface. No model training or fine-tuning is performed.

For the hosted replication, we run the same ToolSandbox suites with the DeepSeek API
model identifiers \texttt{deepseek-chat} and \texttt{deepseek-reasoner} through the same
OpenAI-compatible client. API keys are loaded from environment variables and are not
written to command lines or result files.

For AppWorld, we use two settings. First, we run an isolated \texttt{pctu-appworld}
environment on a targeted public stateful slice covering 72 tasks across twenty-four
registered AppWorld intent machines; some hosted DeepSeek rows are combined from an
earlier 66-task run plus the final six-task Spotify-family expansion. Second, we run
ICVE on the complete local official AppWorld 0.2.0 \texttt{dev.txt} split in a separate
\texttt{pctu-appworld-agents} environment and evaluate with AppWorld's packaged
evaluator. The deterministic path uses typed compilers only; the real-LLM paths use
local Qwen2.5-3B, DeepSeek-chat, or DeepSeek-reasoner to produce a typed intent frame
and then execute that frame through the same runtime-checked executor. AppWorld handlers
interact only through the live \texttt{apis} object; the runtime policy blocks code that
references task files, ground truth, compiled solutions, or oracle data.

Metrics are task success, ToolSandbox similarity, invalid tool calls per task, unsafe
state changes per task, verifier rejections, runtime repair actions, model calls, and a
token proxy computed from local model usage. The kill criterion is not only safety: a
candidate must reduce invalid calls and unsafe state changes without letting repair cost
erase task-success or token-efficiency gains.
For AppWorld, we report evaluator success, unsafe state-test failures, failed API repair
attempts, API calls, model calls, and model-reported token usage.

We use Wilson 95\% intervals over the fixed episode counts in each main suite; the
complete interval table is archived in the supplement. Invalid and unsafe counts are
reported as per-episode means on the same fixed slices.

The experiments answer three questions: whether ICVE reduces invalid or unsafe state
changes, whether the same boundary transfers across models and environments, and whether
typed machines plus abstention are necessary beyond intent recognition.
The anonymous supplement and 4open.science mirror include source code, task-id lists,
evaluator summaries, and commands for reproducing the reported rows, while excluding
credentials, raw logs, databases, and downloaded bundles.

\paragraph{Machine construction protocol.}
Static machines are developed from API signatures, public task instructions, and
ordinary runtime error categories, not from ground-truth answers, private database state,
compiled solutions, or successful trajectories. The AppWorld runner enforces this
discipline at execution time by loading only the public instruction and routing all state
access through the live \texttt{apis} object. Machines are registered at the intent-family
level: a compiler emits reusable slots and a handler must either operate through public
APIs or return an unsupported no-action outcome. We audit this boundary with per-machine
coverage/LOC tables, the full local \texttt{test\_normal.txt} coverage diagnostic, and a
static public-instruction compile audit over \texttt{test\_normal.txt} and
\texttt{test\_challenge.txt}.

\paragraph{Baselines.}
We compare against ReAct \citep{yao2022react}, a plain JSON tool-use loop with no runtime
intent compiler; ReAct+guardrail (PCTU), a stronger proof-carrying baseline that keeps
the model in charge of the action loop but requires each mutating call to carry a schema,
evidence, preconditions, and postconditions checked by a runtime verifier; and ICVE
ablations that remove the typed DSL or the abstention verifier.

\paragraph{Implementation audit.}
We checked the large ICVE--ReAct gaps against three common experimental artifacts. First,
the ToolSandbox rows use the same scenario resolver, available tools, model endpoint,
temperature, message budget, and evaluator for all methods; the ReAct agent receives the
same tool signatures as ICVE, and its tool calls are executed through the same
environment-facing code path. Second, AppWorld comparisons are evaluated by AppWorld's
packaged evaluator on the same task files where applicable; the official dev57 ReAct row
uses the upstream \texttt{simplified\_react\_code\_agent} with its default 50-step
budget. Third, the AppWorld runner now loads only the public task instruction before
execution, and both runtime and prompting policies block references to task files,
ground-truth data, compiled solutions, or oracle fields. These checks do not make the
comparison universal; they rule out known wrapper bugs, evaluator mismatch, and hidden
answer access. The remaining threat is coverage: ICVE is strong where an accurate machine
is registered, and abstains or falls back outside that coverage.

\section{Results}

\subsection{MiniStore Diagnostic}

MiniStore is a small local stateful environment used as a diagnostic rather than as the
headline benchmark. It tests whether simply adding a schema/proof/verifier guardrail to
ReAct is enough.

\begin{table}[h]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lrrrrrr}
\hline
method & succ. & goal & invalid & unsafe & LLM & tokens \\
\hline
ReAct & 0.375 & 1.000 & 1.125 & 1.125 & 5.500 & 2477.750 \\
ReAct+guardrail & 0.250 & 0.250 & 0.000 & 0.000 & 7.250 & 3169.375 \\
\icve & 1.000 & 1.000 & 0.000 & 0.000 & 2.250 & 603.875 \\
\hline
\end{tabular}
\caption{MiniStore real-LLM diagnostic with Qwen2.5-3B. ReAct+guardrail (PCTU)
suppresses invalid and unsafe actions, but still leaves the model to author proof-bearing
actions and pays too much proof/repair overhead.}
\label{tab:ministore}
\end{table}

\subsection{Public ToolSandbox}

The strongest evidence is on ToolSandbox \citep{lu2024toolsandbox}. Table
\ref{tab:toolsandbox-main} reports the primary Qwen2.5-3B results, and
Figure~\ref{fig:toolsandbox-deep} summarizes the scale and failure-mode effects.

\begin{figure}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{2pt}
\newcommand{\ratebar}[2]{\fcolorbox{black!15}{#1}{\rule{#2}{0.12cm}}}
\newcommand{\stackseg}[2]{{\color{#1}\rule{#2}{0.15cm}}}
\begin{tabular}{@{}lccc@{}}
\multicolumn{4}{@{}l}{\textbf{(a) Success across model scales}}\\[-0.15em]
\toprule
model & ReAct & \icve & closed gap \\
\midrule
0.5B & \ratebar{red!22}{0.05cm} .03 & \ratebar{blue!28}{1.35cm} 1.00 & +.97 \\
3B & \ratebar{red!22}{0.10cm} .07 & \ratebar{blue!28}{1.35cm} 1.00 & +.93 \\
7B & \ratebar{red!22}{0.20cm} .13 & \ratebar{blue!28}{1.35cm} 1.00 & +.87 \\
DeepSeek & \ratebar{red!22}{0.18cm} .12 & \ratebar{blue!28}{1.35cm} 1.00 & +.88 \\
\bottomrule
\end{tabular}
\vspace{0.25em}

\begin{tabular}{@{}lcl@{}}
\multicolumn{3}{@{}l}{\textbf{(b) Insuff.-info terminations}}\\[-0.15em]
\toprule
method & distribution & mix \\
\midrule
ReAct & \stackseg{blue!65}{1.15cm}\stackseg{red!70}{0.99cm}\stackseg{gray!55}{0.16cm}
& .50 / .43 / .07 \\
\icve & \stackseg{blue!65}{2.30cm}
& 1.00 / .00 / .00 \\
\bottomrule
\end{tabular}
\caption{Deep ToolSandbox analysis. ICVE closes the success gap across weak, main, and
hosted model scales and changes insufficient-information failures by eliminating unsafe
state transitions. Blue denotes success, red unsafe mutation, and gray other invalid or
incomplete termination.}
\label{fig:toolsandbox-deep}
\end{figure}

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llrrrrrrr}
\toprule
suite & method & n & succ. & invalid & unsafe & repair & LLM & tokens \\
\midrule
single-turn & ReAct & 73 & 0.0685 & 2.1507 & 0.0000 & 0.0000 & 11.8219 & 11048.0 \\
single-turn & ICVE no DSL & 73 & 0.4932 & 0.3288 & 0.0000 & 0.8082 & 3.9726 & 3307.9 \\
single-turn & \icve & 73 & 1.0000 & 0.0000 & 0.0000 & 2.4658 & 0.1233 & 32.7 \\
insufficient-info & ReAct & 28 & 0.5000 & 2.1786 & 0.4286 & 0.0000 & 11.7500 & 9896.4 \\
insufficient-info & ICVE no abstention & 28 & 0.5000 & 1.5714 & 0.4286 & 0.3929 & 7.2857 & 6612.1 \\
insufficient-info & \icve & 28 & 1.0000 & 0.0000 & 0.0000 & 1.5714 & 0.0000 & 0.0 \\
\bottomrule
\end{tabular}
\caption{Main public ToolSandbox results with Qwen2.5-3B. Metrics after success are
per-episode means. ICVE performs deterministic repair actions but still uses far fewer
model calls and tokens; the typed DSL is necessary on single-turn tasks, and the
abstention verifier is necessary on insufficient-information tasks.}
\label{tab:toolsandbox-main}
\end{table*}

The Wilson 95\% confidence intervals for the primary success rates are separated from the
mean-count columns: on the 73-episode single-turn suite, ICVE is 1.0000
[$0.9500, 1.0000$] while ReAct is 0.0685 [$0.0296, 0.1505$]; on the 28-episode
insufficient-information suite, ICVE is 1.0000 [$0.8794, 1.0000$] while ReAct and the
no-abstention ablation are both 0.5000 [$0.3263, 0.6737$].

\begin{table}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{tabular}{lrrrrr}
\toprule
method & succ. & invalid & unsafe & reject & LLM \\
\midrule
ReAct & .321 & 1.214 & .607 & .000 & 7.500 \\
ReAct+guardrail & .750 & 1.036 & .179 & 2.214 & 6.786 \\
\icve & 1.00 & .00 & .00 & .00 & .00 \\
\bottomrule
\end{tabular}
\caption{DeepSeek-chat ToolSandbox guardrail diagnostic on the full 28-scenario
insufficient-information suite. ReAct+guardrail (PCTU) improves safety by rejecting
model-authored contracts, but leaves proof/action authoring in the model loop and retains
nonzero invalid/unsafe calls and model cost.}
\label{tab:toolsandbox-guardrail}
\end{table}

Table~\ref{tab:toolsandbox-guardrail} extends the proof-carrying baseline beyond the
MiniStore diagnostic to the full DeepSeek-chat insufficient-information suite: guardrails
reduce but do not eliminate unsafe or invalid termination, whereas ICVE removes the
mutating loop from the model for covered intents and executes checked transitions without
agent LLM calls.

\paragraph{Dynamic-synthesis probe.}
To test whether ICVE can add a machine without pre-registering that family, we disable
the static registry and enable the affordance-induction prototype described in
Section~2.2. On five official ToolSandbox setting scenarios
(\texttt{cellular\_off}, \texttt{wifi\_off}, and three low-battery repair cases), the
runtime synthesizes one setting machine per task from the available API signatures,
passes shadow-mode and counterexample checks, and reaches 5/5 success with zero LLM
calls. We also run a held-out synthetic-affordance diagnostic with no static machines:
five previously unseen APIs (\texttt{bluetooth}, \texttt{dark\_mode},
\texttt{privacy\_mode}, \texttt{roaming\_data}, and \texttt{auto\_sync}) are induced and
mutated correctly, while an invalid \texttt{focus\_mode} setter and an unrelated stock
query are rejected. The diagnostic is intentionally limited to regular boolean setting
APIs; it does not claim automatic synthesis for AppWorld purchase, Gmail, or
vacation-expense machines.

\subsection{Model Scale}

The effect is not a Qwen2.5-3B-only artifact. Table \ref{tab:model-scale} summarizes weak
and stronger local model runs. The 7B result uses a local 4-bit quantized model and should
not be interpreted as frontier-model coverage.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lllrrrrr}
\toprule
model & suite & method & n & succ. & invalid & unsafe & tokens \\
\midrule
0.5B & single-turn & ReAct & 30 & 0.0333 & 5.6667 & 0.0000 & 8054.5 \\
0.5B & single-turn & \icve & 30 & 1.0000 & 0.0000 & 0.0000 & 54.3 \\
0.5B & insufficient-info & ReAct & 28 & 0.6071 & 3.2500 & 0.3214 & 8367.3 \\
0.5B & insufficient-info & \icve & 28 & 1.0000 & 0.0000 & 0.0000 & 0.0 \\
7B-4bit & single-turn & ReAct & 30 & 0.1333 & 0.4333 & 0.0000 & 3483.5 \\
7B-4bit & single-turn & \icve & 30 & 1.0000 & 0.0000 & 0.0000 & 34.6 \\
7B-4bit & insufficient-info & ReAct & 28 & 0.4286 & 1.2143 & 0.3929 & 4894.4 \\
7B-4bit & insufficient-info & \icve & 28 & 1.0000 & 0.0000 & 0.0000 & 0.0 \\
\bottomrule
\end{tabular}
\caption{ICVE preserves the zero-invalid, zero-unsafe pattern across weak and stronger
local Qwen2.5 scales; the primary Qwen2.5-3B results appear in Table
\ref{tab:toolsandbox-main}.}
\label{tab:model-scale}
\end{table*}

\subsection{Cross-Family Diagnostic}

We also ran a small non-Qwen diagnostic with Phi-3-mini: on 10 core ToolSandbox
scenarios, ReAct reaches 0/10 success with 1.8 invalid calls per task, while ICVE reaches
10/10 with zero invalid/unsafe calls; on 10 insufficient-information scenarios, ReAct
reaches 6/10 with 1.7 invalid and 0.3 unsafe calls per task, the no-abstention ablation
keeps 6/10 and restores unsafe attempts, and ICVE reaches 10/10 with zero
invalid/unsafe calls. This is not frontier-model evidence, but it addresses the narrow
concern that the result is only a Qwen-family artifact.

\subsection{Hosted OpenAI-Compatible Replication}

We also replicate the primary ToolSandbox suites with the hosted DeepSeek-chat endpoint.
Table \ref{tab:deepseek-replication} shows the same qualitative pattern: ICVE eliminates
invalid and unsafe outcomes on both covered suites, while ReAct remains substantially less
successful and more expensive. The no-DSL and
no-abstention ablations again expose the two mechanisms: typed compilation is needed on
single-turn stateful tasks, and abstention verification is needed when the user omits
required information.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llrrrrr}
\toprule
suite & method & n & succ. & invalid & unsafe & tokens \\
\midrule
single-turn & ReAct & 73 & 0.1233 & 0.5342 & 0.0000 & 4397.1 \\
single-turn & ICVE no DSL & 73 & 0.5479 & 0.0548 & 0.0000 & 1258.6 \\
single-turn & \icve & 73 & 1.0000 & 0.0000 & 0.0000 & 13.6 \\
insufficient-info & ReAct & 28 & 0.3214 & 1.2143 & 0.6071 & 7299.6 \\
insufficient-info & ICVE no abstention & 28 & 0.5000 & 0.6786 & 0.4286 & 3234.4 \\
insufficient-info & \icve & 28 & 1.0000 & 0.0000 & 0.0000 & 0.0 \\
\bottomrule
\end{tabular}
\caption{Hosted DeepSeek-chat ToolSandbox replication. Metrics after success are
per-episode means; full outputs are archived in the artifact manifest.}
\label{tab:deepseek-replication}
\end{table*}

We also ran DeepSeek-reasoner on the same two suites; both full runs passed the kill
criteria, and the recorded summaries are listed in the artifact manifest.

\subsection{AppWorld}

The same runtime-checked executor transfers to AppWorld on a targeted public state-changing
slice. The active rows cover 72 tasks across twenty-four registered AppWorld
intent machines spanning phone, Venmo, Gmail, Simple Note, Spotify, Amazon, alarms,
Bucket List, and cross-app sharing tasks.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrr}
\toprule
method & n & succ. & inv./repair & unsafe & LLM & tokens \\
\midrule
deterministic compiler & 72 & 72/72 & 0.0694 & 0.0000 & 0.0000 & 180.5 \\
Qwen2.5-3B intent & 72 & 72/72 & 0.0694 & 0.0000 & 1.0000 & 3220.3 \\
DeepSeek-chat intent & 72 & 72/72 & 0.0694 & 0.0000 & 1.0000 & 3352.2 \\
DeepSeek-reasoner intent & 72 & 72/72 & 0.0694 & 0.0000 & 1.0000 & 3462.3 \\
Qwen direct code & 72 & 0/72 & 1.0000 & 0.0000 & 1.0000 & 1544.0 \\
Qwen typed intent/code & 72 & 0/72 & 1.0000 & 0.0000 & 2.0000 & 4674.8 \\
Qwen code repair & 72 & 0/72 & 2.7083 & 0.0000 & 2.8889 & 5342.8 \\
Qwen multi-step code & 72 & 0/72 & 3.7500 & 0.0000 & 4.3472 & 9031.7 \\
DeepSeek direct code & 72 & 4/72 & 0.8611 & 0.0139 & 1.0000 & 2028.9 \\
DeepSeek typed intent/code & 72 & 7/72 & 0.7917 & 0.0139 & 2.0000 & 5478.9 \\
DeepSeek code repair & 72 & 21/72 & 1.5556 & 0.0417 & 2.3472 & 5898.8 \\
DeepSeek multi-step code & 72 & 20/72 & 1.2222 & 0.0000 & 4.3056 & 12128.6 \\
\bottomrule
\end{tabular}
\caption{Targeted public AppWorld slice. Metrics after success are per-task means. All
rows use the same 72-task dataset; hosted DeepSeek rows are combined from incrementally
flushed shards where needed. AppWorld's packaged evaluator confirms the corresponding
task-completion counts: 72/72 for ICVE runtime rows, 0/72 for local Qwen code baselines,
4/72 for DeepSeek direct code, 7/72 for DeepSeek typed intent/code, 21/72 for DeepSeek
repair, and 20/72 for DeepSeek multi-step code.}
\label{tab:appworld-slice}
\end{table*}

The AppWorld result separates intent recognition from checked execution. Typed-intent
code gives the model the same frame but still asks it to write the mutation code; it
reaches 0/72 with Qwen2.5-3B and 7/72 with DeepSeek-chat. Direct code and multi-step
code-observation variants likewise fail or become expensive, while DeepSeek repair reaches
21/72 but introduces unsafe changes. The error pattern is consistent: models often know
the requested operation but fail at grounded execution. One pre-verifier DeepSeek-chat
intent extraction produced \texttt{archive} for a delete instruction; the
instruction-aware slot verifier rewrites the frame before execution. ICVE's residual
repair cost is
concentrated in Venmo transfers, where public metadata reveals card expiry but not hidden
balances, so the runtime probes non-expired cards and then stops after success.

We then moved the ICVE binding to the complete local official AppWorld 0.2.0
\texttt{dev.txt} split. This adds the remaining official dev task families as registered
intent machines, including navigation-based Spotify artist lookup, Venmo social-feed
aggregation, file-prefix moves, current-artist follower lookup, and Simple Note markdown
export. Table~\ref{tab:appworld-dev57} reports AppWorld's packaged evaluator on all 57
dev tasks. Both deterministic \icve and \icve with DeepSeek-chat intent extraction
complete all 57 local dev tasks with 100.0 task-goal / 100.0 scenario-goal completion,
while the official DeepSeek ReAct-code agent reaches 79.0 / 73.7 and 45/57. The residual
0.0351 invalid API attempts per task are two Venmo payment-card probes: public card
metadata exposes expiry but not balance, so the runtime tries one non-expired
insufficient-balance card and then succeeds with the next public card. We do not use the
private admin balance API to remove this residual repair cost.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrr}
\toprule
method & n & task & scen. & invalid & unsafe & LLM \\
\midrule
deterministic compiler & 57 & 100.0 & 100.0 & 0.0351 & 0.0000 & 0.0000 \\
DeepSeek-chat intent & 57 & 100.0 & 100.0 & 0.0351 & 0.0000 & 1.0000 \\
official DeepSeek ReAct-code & 57 & 79.0 & 73.7 & -- & -- & 15.4211 \\
\bottomrule
\end{tabular}
\caption{Complete local official AppWorld 0.2.0 dev split for ICVE rows. Task and
scenario columns are packaged-evaluator percentages; invalid, unsafe, and LLM
are per-task means from the local runtime logs for ICVE and official runner logs for
ReAct-code. The official ReAct-code row uses 6.45M tokens total.}
\label{tab:appworld-dev57}
\end{table*}

The official DeepSeek ReAct-code dev57 run uses 879 LLM calls and 6.45M tokens. These are
same-split local official-dev comparisons, not held-out test or leaderboard submissions.

As a held-out smoke beyond the official dev split, \texttt{test\_normal} initially
exposed zero coverage on the first 12 tasks; after registering four typed machines, ICVE
reached 12/12. Expanding to the first 120 tasks gives 117/120 overall and 117/117
supported success, with the three then-unsupported trip-note debt-settlement tasks
abstained. After adding one general machine for Simple Note trip-debt ledgers to private
Venmo payments/requests, the full local file, including its final no-newline task ID,
gives the deterministic profile 168/168 overall, 168/168 supported success, 0 unsafe
changes, and 0 invalid calls. This is a coverage diagnostic, not a public leaderboard
submission.

\paragraph{Coverage and development cost.}
The coverage assumption is auditable rather than implicit. ToolSandbox uses 13 static
machines; AppWorld has 126 registered machines, of which 55 cover all 168
\texttt{test\_normal.txt} tasks (3.05 tasks per used machine). For used AppWorld
machines, the median slots/compiler LOC/handler LOC/total LOC are 1/18/74/94. A static
public-instruction audit compiles 168/168 \texttt{test\_normal} and 141/417
\texttt{test\_challenge} instructions without tools, databases, or ground truth; the
roadmap keeps unsupported Amazon purchase/search (152), Gmail (83), Splitwise (28), Spotify (12), and Venmo (1)
buckets as coverage gaps. Forty-one held-out \texttt{test\_challenge} slices now cover
123 additional tasks across phone, Gmail, file-system, payment-card, Venmo, Spotify, Amazon
shopping, Amazon answer-only, subscription-history, and cross-app workflows, reaching
123/123 success with zero invalid or unsafe calls. These numbers are deliberately reported as
machine-family reuse and coverage accounting, not as a full challenge-set score.

\subsection{Multi-Turn Diagnostic}

On a 28-episode hidden-task multi-turn diagnostic with Qwen2.5-0.5B, ICVE reaches
0.8929 success, 0.9773 mean similarity, 0 invalid calls, and 0 unsafe state changes.
This is an extension result, not the headline claim: a small number of recency and
benchmark-interface near-misses remain.

\section{Analysis}

The empirical pattern is that free-form agents overpay for tasks whose state semantics
are compact but strict. ReAct repeatedly infers evidence needs, calls tools with fragile
arguments, interprets observations, repairs failures, and detects completion. PCTU
rejects bad calls, but still relies on model-authored proof-bearing actions. ICVE moves
these recurring checks into the runtime. Runtime actions can increase, but model calls
and token usage collapse because repeated model-mediated repair disappears.

The size of the improvement is therefore expected on covered tasks rather than evidence
that the base model has learned substantially better reasoning. The covered families have
reusable state semantics: ground an entity, establish read-side evidence, repair a small
set of preconditions, mutate once or in a bounded loop, and stop when the postcondition
holds. ReAct solves that control problem anew on every episode. Its failures concentrate
in invalid JSON or unsupported arguments, missing evidence searches, premature or
repeated mutations, and insufficient-information cases where a text loop guesses instead
of abstaining. ICVE removes those degrees of freedom: the compiler chooses a typed frame,
the handler emits only schema-valid actions over grounded slots, the ledger remembers
observed state and failed calls, the verifier blocks unavailable or unsafe transitions,
and the completion detector prevents duplicate actions. Table~\ref{tab:failure-taxonomy}
summarizes this failure taxonomy.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{p{0.15\textwidth}p{0.20\textwidth}p{0.19\textwidth}p{0.22\textwidth}p{0.15\textwidth}}
\toprule
failure type & ReAct manifestation & guardrail manifestation & ICVE mechanism & evidence position \\
\midrule
Invalid call & Fragile JSON/API arguments and repeated failed tool calls. & Rejects
malformed or ungrounded contracts, then asks the model to retry. & Schema-checked
IntentFrame; deterministic handler emits validated tool arguments. &
Tables~\ref{tab:toolsandbox-main}, \ref{tab:toolsandbox-guardrail}. \\
Missing evidence & Model guesses values or searches ad hoc before mutation. & Requires
evidence fields but still relies on model-authored proof/action pairs. & Ledger-driven
evidence acquisition and grounded slot checks before mutation. &
Section~4.3; AppWorld dev57 and test-normal diagnostics. \\
Precondition uncertainty & Text loop repairs settings or payment state after failures. &
Verifier detects failed preconditions but incurs rejection/repair loops. & Bounded
runtime repair before a mutation, followed by postcondition stopping. &
ToolSandbox repair columns and AppWorld Venmo card probes. \\
Insufficient information & ReAct may mutate despite missing user evidence. & Reduces
unsafe calls, but retains nonzero unsafe/invalid termination and model cost. & Abstention
frame or guarded final action with no model-owned mutation loop. &
Table~\ref{tab:toolsandbox-guardrail}; Fig.~\ref{fig:toolsandbox-deep}. \\
Uncovered intent & Agent may try plausible tools for a similar-looking task. & Coverage
is implicit in prompt/verifier behavior. & No registered machine means no high-risk
mutation; static and executable coverage gaps are reported. & AppWorld coverage audit. \\
\bottomrule
\end{tabular}
\caption{Execution-boundary failure taxonomy. Covered high-risk failures are converted
from model-owned mutation loops into checked transitions, deterministic repair, or safe
abstention.}
\label{tab:failure-taxonomy}
\end{table*}

Ablations support this mechanism. Removing the typed DSL leaves the model with runtime
guards but loses much of the single-turn success and invalid-call control; removing
abstention reproduces the unsafe insufficient-information pattern seen in ReAct. In
AppWorld, giving the model a typed intent but still asking it to author mutation code
does not close the gap, while routing the same frame through the runtime-checked executor
does. The result is a boundary effect: ICVE is clean where an accurate machine covers the
task, but this hand-engineered coverage does not imply broad success on uncovered
families.

The added diagnostics make this boundary falsifiable. On ToolSandbox
insufficient-information tasks, ReAct produces 0.4286 unsafe mutations and 2.1786 invalid
calls per task; ICVE reduces both to zero, while removing abstention restores the unsafe
rate. On AppWorld, the 168-task local \texttt{test\_normal.txt} diagnostic, 141/417 static
\texttt{test\_challenge} compile audit, and 123-task held-out slice collection expose both
coverage and non-coverage. The dynamic diagnostic also induces five boolean-setting
machines from unseen API signatures without model calls.

\section{Related Work}
ReAct-style agents interleave reasoning and action through text \citep{yao2022react};
Reflexion adds verbal self-feedback \citep{shinn2023reflexion}, and Toolformer explores
tool-use learning \citep{schick2023toolformer}. Tool-learning systems and
function-calling datasets scale API selection and invocation
\citep{li2023apibank,shen2024smalltoollearners,qin2024toolllm,patil2024gorilla,
liu2025toolace,dang2025guidedtemplates}, while ToolSandbox, AppWorld, BFCL, tau-bench,
and agent benchmarks evaluate stateful and interactive use
\citep{lu2024toolsandbox,appworld2024,bfcl2024,taubench2024,liu2024agentbench,
zhou2024webarena,xie2024osworld,wang2025dialogtool}. Guardrails, schemas,
proof-carrying prompts, and safety evaluations constrain calls
\citep{ruan2024toolemu,xie2025toolsafety,faghih2025toolpreferences}, but ICVE moves
evidence search, repair, mutation, and stopping into executable intent-level state
machines. Semantic parsing and task-oriented dialogue use intent/slot structure; ICVE
treats frames as substrates for checked execution.

\section{Conclusion}

ICVE reframes reliable stateful tool use as runtime-checked execution: the agent supplies
intent, while a typed runtime owns risky transitions. Across covered ToolSandbox and
AppWorld slices, this boundary reduces invalid calls, unsafe changes, and repair cost;
broader leaderboard coverage remains next.

\section*{Limitations}

The current strongest claim is intentionally narrow. The clean real-LLM evidence is on
ToolSandbox no-distraction single-turn and insufficient-information suites plus targeted
AppWorld transfer, complete local official AppWorld dev57, and local
\texttt{test\_normal.txt} file-level runs. These runs use AppWorld's packaged evaluator
where applicable, but they are not a public AppWorld leaderboard submission. Unsupported
\texttt{test\_challenge} product-search purchase, many Gmail, and vacation-expense tasks
remain blocked as unsupported no-action results rather than forced through similar-looking
machines. Multi-turn coverage is promising but not yet a headline claim.

The explicit, partly hand-engineered registry improves auditability but limits open-domain
coverage. Our affordance-induction prototype shows one conservative extension path: log
unsupported intent frames, induce schemas and handlers only for signature-identifiable
boolean setting APIs, execute candidates first in shadow mode, and promote them only
after runtime invariants and counterexample tests pass. This improves coverage for a
recognizable API family, but it is not open-ended program synthesis. Extending the idea
to AppWorld purchase, Gmail, vacation-expense, or other multi-entity workflows remains future
work; newly proposed machines must be checked before they are trusted to mutate state.

The runtime does not eliminate the need for a language model. It changes where the model
is trusted. ICVE still needs the model for uncovered intents, natural-language
clarification, and domains whose state machines have not been registered. The safety claim
therefore applies to covered high-risk intents, not arbitrary tool execution.

Finally, hosted-model coverage remains narrow: the evidence spans Qwen2.5 scales, a
small Phi-3-mini diagnostic, and hosted DeepSeek replications, but not a broad set of
proprietary frontier models.

\section*{Ethical Considerations}

Reliable stateful tool execution has positive applications in reducing accidental
mutations, repeated tool calls, and unsafe guessing by deployed agents. The same runtime
discipline could also make high-impact automated systems more capable, so the method
should be paired with domain authorization, audit logs, and tool-level access controls.
ICVE is not a substitute for policy enforcement: it checks covered state transitions
against declared machine-readable rules, but cannot guarantee safety for tools or intents
outside the registry.

\bibliography{rave_refs}

\clearpage
\appendix

\section{Claim-to-Evidence Matrix}
\label{app:claim-evidence}

Table~\ref{tab:claim-evidence} summarizes the main claims, the evidence that would make
each claim falsifiable, and the boundary under which the claim should be read. The matrix
is not a substitute for the main experiments; it is a compact audit trail for interpreting
the results without treating covered-task success as open-domain tool-use safety.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{p{0.18\textwidth}p{0.25\textwidth}p{0.35\textwidth}p{0.15\textwidth}}
\toprule
Claim & What would weaken or falsify it & Evidence location & Boundary \\
\midrule
Stateful failures are execution-boundary failures, not only reasoning failures. &
ReAct or ReAct+guardrail would match ICVE's safety and cost while keeping the model-owned
loop. &
ToolSandbox insufficient-info: ReAct+guardrail reaches .750 success but still has .179
unsafe mutations, 1.036 invalid calls, 2.214 verifier rejections, and 6.786 LLM calls per
task; ICVE reaches 1.00 success with zero unsafe, zero invalid, and zero agent LLM calls.
The MiniStore diagnostic shows the same proof/retry cost pattern. &
Covered stateful tool-use families, not arbitrary agents. \\
\addlinespace
ICVE is more than a per-task script bank. &
The registry would have near one machine per benchmark task with no cross-task reuse. &
ToolSandbox uses 13 static machines. AppWorld has 126 registered machines; 55 used
machines cover 168 \texttt{test\_normal.txt} tasks, or 3.05 tasks per used
machine. The artifact lists per-machine slots, compiler LOC, handler LOC, total LOC, and
API namespaces. &
Manual engineering remains part of the current method. \\
\addlinespace
The AppWorld machines are not written from hidden answers. &
Execution would require task answers, private database state, compiled solutions, or
successful trajectories. &
The runner loads only public instructions and accesses state through live
\texttt{apis}; policies block task-file, ground-truth, compiled-solution, and oracle
references. The static coverage audit compiles instructions without starting AppWorld,
executing tools, inspecting databases, or loading ground truth. &
This is an implementation audit, not a proof of independent third-party development. \\
\addlinespace
Typed intent recognition alone does not explain the AppWorld gains. &
Giving the model the same typed frame and asking it to write mutation code would close
the gap. &
On the 72-task AppWorld slice, Qwen typed-intent/code reaches 0/72 and DeepSeek
typed-intent/code reaches 7/72, while ICVE runtime rows reach 72/72 with zero unsafe
changes. &
Targeted AppWorld slice and local evaluator evidence. \\
\addlinespace
Coverage gaps become reported no-action outcomes rather than hidden unsafe mutation. &
Unsupported or similar-looking tasks would be forced through machines and mutate state
unsafely. &
On local \texttt{test\_normal.txt}, ICVE supports and succeeds on 168/168 tasks with
zero unsafe state changes. A static public-instruction audit compiles only 141/417
\texttt{test\_challenge} tasks, and the executable prefix still leaves 9/24 Amazon
product-search purchase tasks unsupported under conservative Amazon machines,
again with zero unsafe
changes. The derived coverage roadmap records the required machine capabilities and
validation gates before these buckets should be claimed as supported. &
Lower success is expected outside registered coverage. \\
\addlinespace
The dynamic path is a bounded coverage extension, not open-ended synthesis. &
The inducer would fail on unseen regular affordances or accept unrelated/invalid APIs. &
With no static machines, five unseen boolean-setting APIs are induced and executed
correctly; an invalid \texttt{focus\_mode} setter and an unrelated stock query are
rejected. Promotion requires shadow-mode execution, invariant checks, and counterexample
tests. &
Regular boolean setting APIs only. \\
\addlinespace
The pattern is not tied to one local model family. &
The effect would disappear outside Qwen2.5-3B. &
The paper reports Qwen2.5-0.5B and 7B-4bit scale checks, Phi-3-mini diagnostics, and
hosted DeepSeek-chat / DeepSeek-reasoner ToolSandbox replications. &
Not broad proprietary frontier-model coverage. \\
\bottomrule
\end{tabular}
\caption{Claim-to-evidence matrix. Each row states how the claim should be interpreted,
where the supporting evidence is reported, and what boundary prevents overclaiming.}
\label{tab:claim-evidence}
\end{table*}

\clearpage
\section{Artifact and Reproducibility Map}
\label{app:artifact-map}

The anonymous supplement is organized so that each reported result has a source-code
entry point and a recorded summary directory. The main files are:
\begin{itemize}[leftmargin=*]
\item The runtime and DSL live under \texttt{src/pctu\_pilot/}; the key modules are the
typed DSL, registry runtime, ToolSandbox binding, and AppWorld binding.
\item Public runners live under \texttt{experiments/}; they cover ToolSandbox, AppWorld,
dynamic setting-machine induction, and held-out boolean-affordance induction.
\item Static AppWorld coverage results live in \texttt{results/appworld\_static\_coverage};
they audit public-instruction compile coverage for local \texttt{test\_normal.txt} and
\texttt{test\_challenge.txt} without executing tools or loading ground truth, and include
\texttt{coverage\_roadmap.csv} for unsupported buckets.
\item Review-strengthening summaries live in the dated \texttt{icve\_review\_strengthening}
result directory; it contains \texttt{summary.json} and the per-machine development-cost
table.
\item The full artifact index is the \texttt{artifact\_manifest\_rave2.md} file under
\texttt{paper/}; it maps paper tables to result directories and records the scope
boundaries used in the paper.
\end{itemize}

The supplement builder and package verifier are both in \texttt{experiments/}. The
verifier checks upload hashes, PDF section placement, supplement text hygiene, redaction
of raw model-output preview fields, absence of raw trajectories, and freshness of the
local transfer sheets. These checks do not prove the scientific claims by themselves;
they verify that the submitted artifacts match the paper and that the reported evidence
is traceable.

\end{document}
